Continuous time

### 2023-09-20 continuous-time-draft.qmd
small fixes

### Fit model
Reset duration within each session
```{r}
dat <- dat |>
    mutate(
        ps = interaction(pid, session, sep = "_")
    ) |>
    filter(prompt == "Wellbeing") |>
    mutate(
        duration = duration - duration[1],
        hours = duration / 60,
        .by = c(pid, session)
    )
write_rds(dat, "data/ct-dev-data.rds")
```

@@

## YOLO model with random slopes
Here I add a random slope to each piece of the spline, i.e. 4 random slopes at the participant level. The model is fit on a random subset of 1000 pids.
```{r}
#| cold-fold: show

# include 1000 participants
set.seed(1337)
pids <- sample(unique(dat$pid), 1000)
tmp <- filter(dat, pid %in% pids)
# so predict won't complain
tmp <- tmp |>
    mutate(
        pid = as.integer(pid),
        ps = as.integer(ps)
    )
# same spline but different parameterization
# usually less estimation problems
spline <- ns(
    dat$hours,
    knots = c(0.12, 0.4, 1),
    Boundary.knots = c(0.016, 3)
)
colnames(spline) <- paste0("hours", 0:3)
tmp <- cbind(dat, spline)
path <- "models/lmm-fit3.Rds"
if (file.exists(path)) {
    fit3 <- read_rds(path)
} else {
    fit3 <- lmer(
        mood ~ hours0 + hours1 + hours2 + hours3 +
         (1 | ps) + 
         (1 + hours0 + hours1 + hours2 + hours3 | pid),
        data = tmp
    )
    write_rds(fit3, path)
}

@@

```{r}
get_trend <- function(fit, data, re_int = 0, re_s0 = 0, re_s1 = 0, re_s2 = 0, re_s3 = 0) {
    hours <- seq(0, 2, length.out = 100)
    x <- ns(
        hours,
        knots = c(0.12, 0.4, 1),
        Boundary.knots = c(0.016, 3)
    )
    b0 <- fixef(fit)[1] + re_int
    b <- fixef(fit)[-1]
    y <- b0 + x[, 1] * (b[1] + re_s0) + x[, 2] * (b[2] + re_s1) + x[, 3] * (b[3] + re_s2) + x[, 4] * (b[4] + re_s3)
    data.frame(
        x = hours,
        y = c(y),
        re_int
    )
}
re <- as.data.frame(VarCorr(fit3))
pid_int <- re[2, "sdcor"]
pid_slope0 <- re[3, "sdcor"]
pid_slope1 <- re[4, "sdcor"]
pid_slope2 <- re[5, "sdcor"]
pid_slope3 <- re[6, "sdcor"]
pid_cor0 <- re[7, "sdcor"]
pid_cor1 <- re[8, "sdcor"]
pid_cor2 <- re[9, "sdcor"]
pid_cor3 <- re[10, "sdcor"]
sd1 <- 1
sd2 <- 2
rbind(
    get_trend(fit3, data = tmp),
    get_trend(
        fit3,
        data = tmp,
        -sd1 * pid_int,
        -sd1 * pid_cor0 * pid_slope0,
        -sd1 * pid_cor1 * pid_slope1,
        -sd1 * pid_cor2 * pid_slope2,
        -sd1 * pid_cor3 * pid_slope3
    ),
    get_trend(
        fit3,
        data = tmp,
        -sd2 * pid_int,
        -sd2 * pid_cor0 * pid_slope0,
        -sd2 * pid_cor1 * pid_slope1,
        -sd2 * pid_cor2 * pid_slope2,
        -sd2 * pid_cor3 * pid_slope3
    )
) |>
    ggplot(
        aes(
            x,
            y,
            color = factor(re_int),
            group = factor(re_int)
        )
    ) +
    geom_line() +
    labs(x = "Session duration (hours)")
```

@@

Zoom in on pid 9263.
```{r}
ggplot(
    data = filter(pred, pid == 9263),
    aes(
        x = hours,
        y = yhat,
        group = ps
    )
) +
    geom_line(
        aes(y = mood),
        linetype = "dashed",
        alpha = 0.33
    ) +
    geom_line(
        color = "blue",
        linewidth = 2
    ) +
    geom_point(
        aes(y = mood)
    ) +
    facet_wrap(~session)
```

@@

```{r}
pred <- fit3@frame
pred$yhat <- predict(fit3)
pred |>
    # compare the last and first value
    dplyr::summarize(change = last(yhat) - yhat[1], .by = c(pid, ps)) |>
    # average over sessions within participants
    dplyr::summarize(change = mean(change), .by = pid) |>
    # average over participants
    dplyr::summarize(mean(change), sd(change))
```

We can also plot the distribution of the participants average change during sessions.
```{r}
pred |>
    dplyr::summarize(change = last(yhat) - yhat[1], .by = c(pid, ps)) |>
    dplyr::summarize(change = mean(change), .by = pid) |>
    ggplot(aes(change)) +
    geom_histogram()
```

They really don't change much. Let's look at a participant out in the tails.
```{r}
pred |>
    dplyr::summarize(change = last(yhat) - yhat[1], .by = c(pid, ps)) |>
    filter(change > 0.1) |>
    head()
```

Zoom in on pid 23.
```{r}
pred2 <- filter(tmp, pid == 23)
pred2$yhat <- predict(fit3, newdata = pred2)
ggplot(
    data = pred2,
    aes(
        x = hours,
        y = yhat,
        group = ps
    )
) +
    geom_line(
        aes(y = mood),
        linetype = "dashed",
        alpha = 0.33
    ) +
    geom_line(
        color = "blue",
        linewidth = 2
    ) +
    geom_point(
        aes(y = mood)
    ) +
    facet_wrap(~session)
```

@@

```{r}
#| code-fold: show
library(brms)
options(
    mc.cores = 8,
    brms.backend = "cmdstanr"
)
fit_b0 <- brm(
    mood ~ hours0 + hours1 + hours2 + hours3 + (1 | ps) + (1 | pid),
    data = tmp,
    #threads = 2,
    chains = 4,
    iter = 500,
    backend = "rstan",
    file = "models/brm-pre-post-ct0"
)
pp_check(fit_b0)
```

Censoring
```{r}
#| code-fold: show
tmp <- tmp |>
    mutate(
        cl = case_when(
            mood == 0 ~ "left",
            mood == 1 ~ "right",
            TRUE ~ "none"
        )
    )
fit_b1 <- brm(
    mood | cens(cl) ~ hours0 + hours1 + hours2 + hours3 + (1 | ps) + (1 | pid),
    data = tmp,
    threads = 2,
    iter = 1000,
    file = "models/brm-pre-post-ct1"
)
pp_check(fit_b1)
```

@@

We'll start with a cubic spline using the same knot placement as used with the LMMs.
```{r}
#| code-fold: show
fit_g_0 <- gam(
    mood ~ s(hours, bs = "cr", k = 5),
    knots = knots,
    method = "REML",
    data = tmp,
)
fit_g_0 <- gam(
    mood ~ s(hours, k = 15),
    #knots = knots,
    method = "REML",
    data = filter(tmp, hours < 5),
)
```

```{r}
plot_gam <- function(fit, t_end = 2) {
    emm0 <- emmeans(
        fit,
        ~hours,
        at = list(hours = seq(0, t_end, length.out = 100))
    )
    p0 <- as.data.frame(emm0) |>
        ggplot(aes(hours, emmean)) +
        geom_line() +
        geom_ribbon(aes(ymin = lower.CL, ymax = upper.CL), alpha = 0.25) +
        labs(x = "Session duration (hours)")
    p0
}
plot_gam(fit_g_0) + lims(y = c(0,1))
```

@@

```{r}
#| code-fold: show
library(gamm4)
fit_g_1_lme4 <- gamm4(
    mood ~ s(hours, bs = "cr", k = 5),
    knots = knots,
    data = tmp,
    random = ~ (1 | pid) + (1 | ps)
)
```

@@

```{r}
#| code-fold: show

>>>>

### Fit model
Reset duration within each session
```{r}
dat <- dat |>
    mutate(
        duration = duration - duration[1],
        hours = duration / 60,
        .by = c(pid, session)
    ) |> mutate(
        ps = interaction(pid, session, sep = "_")
    ) |>
    filter(prompt == "Wellbeing")

dat |> 
    summarize(
        int = duration[1], .by = ps
        ) |> 
    summarize(
        median(int),
        mean(int),
        sd(int)
    )

write_rds(dat, "data/ct-dev-data.rds")
```

@@

## YOLO model with random slopes
Here I add a random slope to each piece of the spline, i.e. 4 random slopes at the participant level. The model is fit on a random subset of 1000 pids.
```{r}
#| cold-fold: show
set.seed(1337)
# same spline but different parameterization
# usually less estimation problems
spline <- ns(
    dat$hours,
    knots = c(0.12, 0.4, 1),
    Boundary.knots = c(0.016, 3)
)
colnames(spline) <- paste0("hours", 0:3)
tmp <- cbind(dat, spline) |> 
    filter(hours <= 3)
path <- "models/lmm-fit3.Rds"
if (file.exists(path)) {
    fit3 <- read_rds(path)
} else {
    fit3 <- lmer(
        mood ~ hours0 + hours1 + hours2 + hours3 +
         (1 | ps) + 
         (1 + hours0 + hours1 + hours2 + hours3 | pid),
        data = tmp
    )
    write_rds(fit3, path)
}

@@

```{r}
get_trend <- function(fit, data, re_int = 0, re_s0 = 0, re_s1 = 0, re_s2 = 0, re_s3 = 0) {
    hours <- seq(0, 3, length.out = 100)
    x <- ns(
        hours,
        knots = c(0.12, 0.4, 1),
        Boundary.knots = c(0.016, 3)
    )
    b0 <- fixef(fit)[1] + re_int
    b <- fixef(fit)[-1]
    y <- b0 + x[, 1] * (b[1] + re_s0) + x[, 2] * (b[2] + re_s1) + x[, 3] * (b[3] + re_s2) + x[, 4] * (b[4] + re_s3)
    data.frame(
        x = hours,
        y = c(y),
        re_int
    )
}
re <- as.data.frame(VarCorr(fit3))
pid_int <- re[2, "sdcor"]
pid_slope0 <- re[3, "sdcor"]
pid_slope1 <- re[4, "sdcor"]
pid_slope2 <- re[5, "sdcor"]
pid_slope3 <- re[6, "sdcor"]
pid_cor0 <- re[7, "sdcor"]
pid_cor1 <- re[8, "sdcor"]
pid_cor2 <- re[9, "sdcor"]
pid_cor3 <- re[10, "sdcor"]
sd1 <- 1
sd2 <- 2
rbind(
    get_trend(fit3, data = tmp),
    get_trend(
        fit3,
        data = tmp,
        -sd1 * pid_int,
        -sd1 * pid_cor0 * pid_slope0,
        -sd1 * pid_cor1 * pid_slope1,
        -sd1 * pid_cor2 * pid_slope2,
        -sd1 * pid_cor3 * pid_slope3
    ),
    get_trend(
        fit3,
        data = tmp,
        -sd2 * pid_int,
        -sd2 * pid_cor0 * pid_slope0,
        -sd2 * pid_cor1 * pid_slope1,
        -sd2 * pid_cor2 * pid_slope2,
        -sd2 * pid_cor3 * pid_slope3
    )
) |>
    ggplot(
        aes(
            x,
            y,
            color = factor(re_int),
            group = factor(re_int)
        )
    ) +
    geom_line() +
    labs(x = "Session duration (hours)")
```

@@

Zoom in on pid 9263.
```{r}
pred <- filter(tmp, pid == "p9263")
pred$yhat <- predict(fit3, newdata = pred)
ggplot(
    data = pred,
    aes(
        x = hours,
        y = yhat,
        group = ps
    )
) +
    geom_line(
        aes(y = mood),
        linetype = "dashed",
        alpha = 0.33
    ) +
    geom_line(
        color = "blue",
        linewidth = 2
    ) +
    geom_point(
        aes(y = mood)
    ) +
    facet_wrap(~session)
```

@@

```{r}
pred <- fit3@frame
pred$yhat <- predict(fit3)
ps_filtered <- pred |> 
    summarise(n = n(), .by = ps) |> 
    filter(n > 1)
change <- pred |>
    # compare the last and first value
    filter(ps %in% ps_filtered$ps) |> 
    summarise(change = last(yhat) - yhat[1], .by = c(pid, ps)) |>
    # average over sessions within participants
    summarise(change = median(change), .by = pid) 
change |> 
    # average over participants
    summarise(mean(change), median(change), sd(change))
```

We can also plot the distribution of the participants average change during sessions.
```{r}
change |> 
    ggplot(aes(change)) +
    geom_histogram()
```

They really don't change much. Let's look at a participant out in the tails.
```{r}
change |> 
    filter(change > 0.1) |>
    head()
```

Zoom in on pid p10057.
```{r}
pred2 <- filter(tmp, pid == "p10057")
pred2$yhat <- predict(fit3, newdata = pred2)
ggplot(
    data = pred2,
    aes(
        x = hours,
        y = yhat,
        group = ps
    )
) +
    geom_line(
        aes(y = mood),
        linetype = "dashed",
        alpha = 0.33
    ) +
    geom_line(
        color = "blue",
        linewidth = 2
    ) +
    geom_point(
        aes(y = mood)
    ) +
    facet_wrap(~session)
```

@@

```{r}
#| eval: false
#| label: brms
#| code-fold: show
library(brms)
options(
    mc.cores = 8,
    brms.backend = "cmdstanr"
)
fit_b0 <- brm(
    mood ~ hours0 + hours1 + hours2 + hours3 + (1 | ps) + (1 | pid),
    data = tmp,
    #threads = 2,
    chains = 4,
    iter = 500,
    backend = "rstan",
    file = "models/brm-pre-post-ct0"
)
pp_check(fit_b0)
```

Censoring
```{r}
#| code-fold: show
#| label: brms-cens
#| eval: false
tmp <- tmp |>
    mutate(
        cl = case_when(
            mood == 0 ~ "left",
            mood == 1 ~ "right",
            TRUE ~ "none"
        )
    )
fit_b1 <- brm(
    mood | cens(cl) ~ hours0 + hours1 + hours2 + hours3 + (1 | ps) + (1 | pid),
    data = tmp,
    threads = 2,
    iter = 1000,
    file = "models/brm-pre-post-ct1"
)
pp_check(fit_b1)
```

@@

We'll start with a cubic spline using the same knot placement as used with the LMMs.
```{r}
#| code-fold: show
fit_g_0 <- gam(
    mood ~ s(hours, bs = "cr", k = 5),
    knots = knots,
    method = "REML",
    data = tmp,
)
```

```{r}
plot_gam <- function(fit, t_end = 2) {
    emm0 <- emmeans(
        fit,
        ~hours,
        at = list(hours = seq(0, t_end, length.out = 100))
    )
    p0 <- as.data.frame(emm0) |>
        ggplot(aes(hours, emmean)) +
        geom_line() +
        geom_ribbon(aes(ymin = lower.CL, ymax = upper.CL), alpha = 0.25) +
        labs(x = "Session duration (hours)")
    p0
}
plot_gam(fit_g_0)
```

@@

```{r}
#| label: gamm4
#| code-fold: show
library(gamm4)
fit_g_1_lme4 <- gamm4(
    mood ~ s(hours, bs = "cr", k = 5),
    knots = knots,
    data = tmp,
    random = ~ (1 | pid) + (1 | ps)
)
```

@@

```{r}
#| label: gamm4-fs
#| code-fold: show
